### 2017-11-15 submission/manuscript.Rmd
complete Nick's comments on discussion, methods, and Declarations

Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the relative risk of either of these two conditions. Even though CRC-associated genera increase the RR of carcinoma, they do not consistently increase the relative risk of adenoma. This information suggests that these specific genera are important in carcinoma genesis but may not be the primary members of the microbial community contributing to the formation of an adenoma. Additionally, our data show that by using the whole community, our models perform better than when only the CRC-associated genera are included. CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions. These observations suggest that small localized changes within the community may be occuring that are important in early disease progression of CRC and that this process may not directly involve CRC-associated genera.

@@

Although there are still questions that need to be answered regarding the microbiota and carcinoma, a clearer framework for their relationship is beginning to develop as to how this occurs. From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 6]. Conversely, the role of the microbiota in adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls. This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma formation.

@@

***Sequence Processing:***
For the majority of studies raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the by searching the respective accession number of the study following website: http://www.ncbi.nlm.nih.gov/Traces/study/. Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap for one study [@ahn_human_2013] and for four studies was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]. Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009]. Where possible, quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing. If it was not possible to use these defaults, the stated quality cut-offs were used instead. Chimeras were identifed and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.

@@

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma. We analyzed the data using linear mixed-effect models and relative risk. $\beta$-diversity was then assessed for each inidividual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched) tissue samples. Where applicable for each study, predictions for adenoma and carcinoma were tested. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. The power of each study was assessed for an effect size ranging from 1% to 30%. An estimated sample n for these effect sizes was also generated based on 80% power.

@@

A detailed and reproducible description of how the data were processed and analyzed for each study can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017. Raw sequences can be downloaded from the SRA in most cases and can be found in the respective studies batch file in the GitHub repository or within the original publication. For instances when sequences are not publicly available, they may be accessed by contacting the corresponding authors from whence the data came.

@@

The authors would like to thank all the study participants who were a part of each of the individual studies uitlized. We would also like to thank each of the study authors for making their data available for use. Finally we would like to thank the members of the Schloss lab for valuable feed back and proof reading during the formulation of this manuscript.

>>>>

Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease [Figure 1-3]. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the relative risk of either of these two conditions. Even though CRC-associated genera increase the RR of carcinoma, they do not consistently increase the relative risk of adenoma. This information suggests that these specific genera are important in carcinoma genesis but may not be the primary members of the microbial community contributing to the formation of an adenoma. Additionally, our data show that by using the whole community, our models perform better than when only the CRC-associated genera are included. CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions. These observations suggest that small localized changes within the community on tissue may be occuring that are important in early disease progression of CRC and that this process may not directly involve CRC-associated genera.

@@

Although there are still questions that need to be answered regarding the microbiota and carcinoma, a clearer framework for their relationship is beginning to develop as to how this occurs. From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 6]. Conversely, the role of the microbiota in adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls. This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out how the microbiota contributes to adenoma formation.

@@

***Sequence Processing:***
For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the by searching the respective accession number of the study following website: http://www.ncbi.nlm.nih.gov/Traces/study/. Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap for one study [@ahn_human_2013] and for four studies was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]. Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009]. Where possible, quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing. If it was not possible to use these defaults, the stated quality cut-offs were used instead. Chimeras were identifed and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.

@@

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma. We analyzed the data using linear mixed-effect models and relative risk. $\beta$-diversity was then assessed for each inidividual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched) tissue samples. Where applicable for each study, predictions for adenoma and carcinoma were tested. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models, the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. The power of each study was assessed for an effect size ranging from 1% to 30%. An estimated sample n for these effect sizes was also generated based on 80% power.

@@

A detailed and reproducible description of how the data were processed and analyzed for each study can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017. Raw sequences can be downloaded from the SRA in most cases and can be found in the respective study batch file in the GitHub repository or within the original publication. For instances when sequences are not publicly available, they may be accessed by contacting the corresponding authors from whence the data came.

@@

The authors would like to thank all the study participants who were a part of each of the individual studies utilized. We would also like to thank each of the study authors for making their data available for use. Finally we would like to thank the members of the Schloss lab for valuable feed back and proof reading during the formulation of this manuscript.

### 2017-11-15 submission/manuscript.tex
complete Nick's comments on discussion, methods, and Declarations

Our study identifies clear differences in diversity both at the
community level and within individual genera that are present in
individuals with CRC versus those without the disease. Although there
was a step-wise decrease in diversity from control to adenoma to
carcinoma, this did not translate into large effect sizes for the
relative risk of either of these two conditions. Even though
CRC-associated genera increase the RR of carcinoma, they do not
consistently increase the relative risk of adenoma. This information
suggests that these specific genera are important in carcinoma genesis
but may not be the primary members of the microbial community
contributing to the formation of an adenoma. Additionally, our data show
that by using the whole community, our models perform better than when
only the CRC-associated genera are included. CRC-associated genera are
clearly important to carcinoma pathogenesis but accounting for the
community in which these microbes exist can drastically increase the
ability of models to make predictions. These observations suggest that
small localized changes within the community may be occuring that are
important in early disease progression of CRC and that this process may
not directly involve CRC-associated genera.

@@

Although there are still questions that need to be answered regarding
the microbiota and carcinoma, a clearer framework for their relationship
is beginning to develop as to how this occurs. From our observations
many changes in carcinoma could easily result in effect sizes that are
30\% or more between the case and control and most studies analyzed have
sufficient power to detect these types of changes {[}Figure 6{]}.
Conversely, the role of the microbiota in adenoma is less clear and part
of the reason this may be is because many studies are not powered
effectively to observe the small changes reported here. None of the
studies analyzed were properly powered to detect a 10\% or lower change
between case and controls. This small effect size range may well be the
scope in which differences consistently occur in adenoma due to the
subtle changes in community that occur between control and adenoma.
Future studies investigating adenoma and the microbiota need to take
these factors into consideration if we are to work out the role of the
microbiota in adenoma formation.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the by searching the respective accession
number of the study following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
for one study {[}21{]} and for four studies was obtained directly from
the authors {[}12,13,18,20{]}. Each study was processed using the mothur
(v1.39.3) software program {[}30{]}. Where possible, quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults, the
stated quality cut-offs were used instead. Chimeras were identifed and
removed using VSEARCH {[}31{]} before \emph{de novo} OTU clustering at
97\% similarity using the OptiClust algorithm {[}32{]} was utilized.

@@

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma. We
analyzed the data using linear mixed-effect models and relative risk.
\(\beta\)-diversity was then assessed for each inidividual study. Next,
four specific CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas})
were assessed for differences in relative risk. We then built Random
Forest models based on all genera or the select CRC-associated genera.
The models were trained on one study then tested on the remaining
studies for every study. The data was split between feces and tissue
samples. Within the tissue groups the data was further divided between
samples from the same individual (matched) and those from different
individuals (unmatched) tissue samples. Where applicable for each study,
predictions for adenoma and carcinoma were tested. This same approach
was then applied at the OTU level with the exception that instead of
testing on the other studies, a 10-fold cross validation was utilized
and 100 different models were created based on random 80/20 splitting of
the data to generate a range of expected AUCs. For OTU based models the
CRC associated genera included all OTUs that had a taxonomic
classification to \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}. The power of each
study was assessed for an effect size ranging from 1\% to 30\%. An
estimated sample n for these effect sizes was also generated based on
80\% power.

@@

A detailed and reproducible description of how the data were processed
and analyzed for each study can be found at
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Raw sequences can be downloaded from the SRA in most cases and can be
found in the respective studies batch file in the GitHub repository or
within the original publication. For instances when sequences are not
publicly available, they may be accessed by contacting the corresponding
authors from whence the data came.

>>>>

Our study identifies clear differences in diversity both at the
community level and within individual genera that are present in
individuals with CRC versus those without the disease {[}Figure 1-3{]}.
Although there was a step-wise decrease in diversity from control to
adenoma to carcinoma, this did not translate into large effect sizes for
the relative risk of either of these two conditions. Even though
CRC-associated genera increase the RR of carcinoma, they do not
consistently increase the relative risk of adenoma. This information
suggests that these specific genera are important in carcinoma genesis
but may not be the primary members of the microbial community
contributing to the formation of an adenoma. Additionally, our data show
that by using the whole community, our models perform better than when
only the CRC-associated genera are included. CRC-associated genera are
clearly important to carcinoma pathogenesis but accounting for the
community in which these microbes exist can drastically increase the
ability of models to make predictions. These observations suggest that
small localized changes within the community on tissue may be occuring
that are important in early disease progression of CRC and that this
process may not directly involve CRC-associated genera.

@@

Although there are still questions that need to be answered regarding
the microbiota and carcinoma, a clearer framework for their relationship
is beginning to develop as to how this occurs. From our observations
many changes in carcinoma could easily result in effect sizes that are
30\% or more between the case and control and most studies analyzed have
sufficient power to detect these types of changes {[}Figure 6{]}.
Conversely, the role of the microbiota in adenoma is less clear and part
of the reason this may be is because many studies are not powered
effectively to observe the small changes reported here. None of the
studies analyzed were properly powered to detect a 10\% or lower change
between case and controls. This small effect size range may well be the
scope in which differences consistently occur in adenoma due to the
subtle changes in community that occur between control and adenoma.
Future studies investigating adenoma and the microbiota need to take
these factors into consideration if we are to work out how the
microbiota contributes to adenoma formation.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the by searching the respective accession
number of the study following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
for one study {[}21{]} and for four studies was obtained directly from
the authors {[}12,13,18,20{]}. Each study was processed using the mothur
(v1.39.3) software program {[}30{]}. Where possible, quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults, the
stated quality cut-offs were used instead. Chimeras were identifed and
removed using VSEARCH {[}31{]} before \emph{de novo} OTU clustering at
97\% similarity using the OptiClust algorithm {[}32{]} was utilized.

@@

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma. We
analyzed the data using linear mixed-effect models and relative risk.
\(\beta\)-diversity was then assessed for each inidividual study. Next,
four specific CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas})
were assessed for differences in relative risk. We then built Random
Forest models based on all genera or the select CRC-associated genera.
The models were trained on one study then tested on the remaining
studies for every study. The data was split between feces and tissue
samples. Within the tissue groups the data was further divided between
samples from the same individual (matched) and those from different
individuals (unmatched) tissue samples. Where applicable for each study,
predictions for adenoma and carcinoma were tested. This same approach
was then applied at the OTU level with the exception that instead of
testing on the other studies, a 10-fold cross validation was utilized
and 100 different models were created based on random 80/20 splitting of
the data to generate a range of expected AUCs. For OTU based models, the
CRC associated genera included all OTUs that had a taxonomic
classification to \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}. The power of each
study was assessed for an effect size ranging from 1\% to 30\%. An
estimated sample n for these effect sizes was also generated based on
80\% power.

@@

A detailed and reproducible description of how the data were processed
and analyzed for each study can be found at
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Raw sequences can be downloaded from the SRA in most cases and can be
found in the respective study batch file in the GitHub repository or
within the original publication. For instances when sequences are not
publicly available, they may be accessed by contacting the corresponding
authors from whence the data came.
